---
title: "White Wine EDA"
author: "George Vince"
date: "August 1, 2017"
---

#Packages used:
```{r echo=TRUE, message=FALSE, warning=FALSE, packages}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(grid)
library(corrplot)
library(RColorBrewer)

```

# Structure of data
```{r echo=TRUE}
wine = read.csv("wineQualityWhites.csv")
names(wine)
str(wine)

# Remove index number, as it's no use
wine$X = NULL
summary(wine)
```
  
# Univariate Plots Section
  Lets take a look at each of the variable histograms

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}

plots <- list()
for (i in 1:length(colnames(wine))){
 
  p <- ggplot(wine, aes_string(colnames(wine[i]))) + 
    geom_histogram()
  plots[[i]] <- p
}


do.call(grid.arrange,c(plots, ncol=4))

wine$quality.fact = factor(wine$quality)

```


It appears that density and pH are normally distributed, with few outliers.
    
Fixed and volatile acidity, sulfur dioxides, sulphates, and alcohol seem to 
be long-tailed.
    
It seems, residual sugar, chlorides, density have extreme outliers.

Lets take a closer look at each of the variables below.


### Quality
```{r echo=FALSE, message=FALSE, warning=FALSE, Quality}
  
  ggplot(wine, aes(quality)) + 
  geom_histogram() +
  scale_x_continuous( limit = c(3,9), breaks = seq(3, 9, by = 1) ) +
  ggtitle('Quality Counts') 
  
    
  table(wine$quality)
```

Here we can see quality ranges from 3 - 9, we don't have much data (if any) for
exceptionally good or bad wines.

Given the ratings and distribution of wine quality, I'll create  another categorical 
variable, classifying the wines as 'bad' (rating 0 to 4), 'average' (rating 5 - 7), 
and 'excellent' (rating 8 to 9).
By splitting the data this way we have roughly the same amount of wines in our 
bad and excellent categories.
```{r echo=TRUE}

wine$rating <- ifelse(wine$quality < 5, 'bad', ifelse(
  wine$quality < 8, 'average', 'excellent'))

wine$rating <- ordered(wine$rating,
                     levels = c('bad', 'average', 'excellent'))

summary(wine$rating)

  
```
### Fixed Acidity

Viewing the histogram it shows that the majority of fixed acidities fall between
5.5 and 8.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot( aes( x= fixed.acidity), data = wine) + 
  geom_histogram() + 
  ggtitle('First plot')

revised <- ggplot( aes( x= fixed.acidity), data = wine) + 
  geom_histogram(binwidth = 0.05) + 
  ggtitle('Fixed Acidity Counts') + 
  scale_x_continuous( limit = c(4,10), breaks = seq(4, 10, by = 0.5) )

summary(wine$fixed.acidity)

grid.arrange(first, revised)
```

### volatile acidity

Volatile acidity produces a long tailed plot with most wines having an acetic 
value between 0.15 - 0.40.  Its stated in the data description that high levels
of acetic acid can cause a vinegary taste, I will see if this results in a poorer
quality rating.

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(volatile.acidity)) + 
  geom_histogram() + 
  ggtitle('First plot')

revised <- ggplot(wine, aes(volatile.acidity)) + 
  geom_histogram(binwidth = 0.005) + 
  scale_x_continuous( limit= c(0.1, 0.6), breaks = seq(0.1, 0.55, by = 0.05) ) + 
  ggtitle('Volitle Counts')

summary(wine$volatile.acidity)

grid.arrange(first, revised)

```

### Citric acid
Small quantities of citric acid can lead to "freshness is wine".
Most values fall between 0.20 - 0.45g/l
Noticed an interesting spike at 0.49g/l


```{r echo=FALSE, message=FALSE, warning=FALSE}
first = ggplot(wine, aes(citric.acid)) +
  geom_histogram() +
  ggtitle('First plot')

revised <- ggplot( aes( x= citric.acid), data = wine) + 
  geom_histogram(binwidth = 0.005) + 
  scale_x_continuous( limit = c(0, 0.8), breaks = seq(0, 0.8, by = 0.05) ) + 
  ggtitle('Citric Acid Counts')

summary(wine$citric.acid)

grid.arrange(first,revised)

```

### Residual Sugar

It seems residual sugar has a large range, but most values fall between 1 - 20g/l. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

first = ggplot(wine, aes(residual.sugar)) +
  geom_histogram()

revised = ggplot(wine, aes(residual.sugar)) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limit = c(1, 20), breaks = seq(1, 20, by = 1)) 
  
  summary(wine$residual.sugar)

  grid.arrange(first,revised)

```

### Chlorides

Chlorides indicate the amount of salt in wine, it seems normally distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(chlorides)) + 
  geom_histogram() + 
  ggtitle('Basic plot')

revised <- ggplot(wine, aes(chlorides)) + 
  geom_histogram(binwidth = 0.001) + 
  scale_x_continuous( limit = c(0, 0.1), breaks = seq(0, 0.1, by = 0.01) ) + 
  ggtitle('Chloride Counts')

summary(wine$chlorides)

grid.arrange(first, revised)
```

### Free sulfur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(free.sulfur.dioxide)) + 
  geom_histogram() + 
  ggtitle('Basic plot')

revised <- ggplot(wine, aes(free.sulfur.dioxide)) + 
  geom_histogram(binwidth = 0.5) + 
  scale_x_continuous( limit = c(0, 80), breaks = seq(0, 100, by = 5) ) + 
  ggtitle('Sulfur dioxide counts')

summary(wine$free.sulfur.dioxide)

grid.arrange(first, revised)


```

According to the data description
"at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and 
taste of wine", so I have created a variable free.sulfur.dioxide.cat as
"noticeable" (free.sulfur.dioxide > 50) and "not noticeable" (free.sulfur.dioxide <= 50).

```{r echo=TRUE}
wine$free.sulfur.dioxide.cat <- ifelse(wine$free.sulfur.dioxide > 50, 
                                       'noticeable', 'not noticeable' )

table(wine$free.sulfur.dioxide.cat)
```

### Total sulfur dioxide

Total sulfure dioxide is made up of free sulfur dioxide and gases given off from
sulphates.  It seems to range from 108 - 167mg/l

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(total.sulfur.dioxide)) + 
  geom_histogram() + 
  ggtitle('Basic plot')

revised <- ggplot(wine, aes(total.sulfur.dioxide)) + 
  geom_histogram(binwidth = 1) + 
  scale_x_continuous( limit = c(10, 250), breaks = seq(10, 250, by = 10) ) + 
  ggtitle('Sulfur Counts')

summary(wine$total.sulfur.dioxide)

grid.arrange(first, revised)

```

### Density

Density is normally distributed with most values falling between 0.9 - 1g /ml;
Density is directly affected by water and alcohol percentage

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(density)) + 
  geom_histogram() + 
  ggtitle('Basic plot')

revised <- ggplot(wine, aes(density)) + 
  geom_histogram(binwidth = 0.0005) + 
  scale_x_continuous( limit = c(0.985, 1.005), breaks = seq(0.985, 1.005, by = 0.005) ) + 
  ggtitle('Density Counts')

summary(wine$density)

grid.arrange(first, revised)
```

### PH
PH Seems to normally distributed and vary from 2.7 - 3.8. This indicates that 
white wines are always acidic.
```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(pH)) + 
  geom_histogram() + 
  ggtitle('PH Count')

summary(wine$pH)

grid.arrange(first)
```

### Sulphates

Sulphates have a mostly normal distribution which is slightly positively 
skewed with few outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(sulphates)) + 
  geom_histogram() + 
  ggtitle('Basic plot')

revised <- ggplot(wine, aes(sulphates)) + 
  geom_histogram(binwidth = 0.01) + 
  scale_x_continuous( limit = c(0.3, 0.8), breaks = seq(0.3, 0.8, by = 0.05) ) + 
  ggtitle('Sulphates Counts')

summary(wine$sulphates)
grid.arrange(first, revised)


```

### Alcohol

Distribution of variable alcohol is positively skewed. Most of the values are 
between 9% and 13%. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
first <- ggplot(wine, aes(alcohol)) + 
  geom_histogram() + 
  ggtitle('Basic plot')

revised <- ggplot(wine, aes(alcohol)) + 
  geom_histogram(binwidth = 0.1) + 
  scale_x_continuous( limit = c(8, 14), breaks = seq(8, 14, by = 0.5) ) + 
  ggtitle('Alcohol Counts')

summary(wine$alcohol)
grid.arrange(first, revised)


```

# Univariate Analysis
  
### What is the structure of your dataset?
  
  My dataset is related to the quality of white wine. The dataset is already tidy
  with observation mentioning the chemical physicochemical of the wine. 
  There are 4898 observations of 13 variables. 
  The variable 'quality' is based on the median rating by 3 wine experts. 
  The other physicochemical  variables are fixed.acidity, volatile.acidity, 
  citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, 
  density, pH, sulphates and alcohol. 
  
  The quality variable can be seen as factor variable with 0 to 10 
  i.e. 11 possible levels, where only 7 levels are present in dataset i.e . 3-9.
  
  It's important to note that only physicochemical variables are available, 
  there is no data on other variables  such as grape type, wine brand, wine 
  selling price etc.
  
### What is/are the main feature(s) of interest in your dataset?
  I'm interested to see if any of the variables produce consistently good, or 
  bad white wine.
  

### Did you create any new variables from existing variables in the dataset?
Binned values in the quality column, providing me with a range of bad, average 
and excellent wines.
I also created a variable to highlight if free bound SO2 levels were noticeable 
in the wine or not.


# Bivariate Plots Section

Lets examine the corrolation matrix

```{r echo=FALSE, Bivariate_Plots}

wine_data <- wine[, sapply(wine, is.numeric)]

corrplot(cor(wine_data), method="number")
```

Key factors that seem to have correlation with quality of wine are density and alcohol.

I'm also going to examine the corolations between density and residual sugar and
the relationship between sulfur dioxides.

pH is also negatively corrolated with fixed acidity, I feel like this is obvious
and doesn't need investigating much further.

## Density
Lets look at the two highest correlating variables with density.
It looks like density is correlated negatively alcohol, and positively with residual 
sugar amounts.  I think intuitively this should be correct as more sugar would result 
in an increase in density, and vice versa for alcohol amounts.

```{r echo=FALSE, message=FALSE, warning=FALSE, density}


p1 = ggplot(wine, aes(density, residual.sugar)) +
  geom_point(alpha=0.1) +
  xlim(0.985, 1.0025)+
  ylim(0, 20)+
  geom_smooth(method = 'lm')+
  ggtitle("Density vs Residual Sugar")

p2 = ggplot(wine, aes(density ,alcohol)) +
  geom_point(alpha=0.1) +
  xlim(0.985, 1.0025) +
  geom_smooth(method = 'lm') +
  ggtitle("Density vs Alcohol")

grid.arrange(p1, p2)

```

## Citric Acid 
I noticed a peak at 0.49g / l, I'm going to see if these wines had any higher quality
then all other wines. Perhaps this is the sweet spot for a fresh tasting wine?

```{r echo=TRUE, message=FALSE, warning=FALSE, alcohol}

wine %>% group_by(citric.acid == 0.49) %>%
    summarise(mean(quality), count = n())

```

Hmm, perhaps not worth investigating this as there is no noticeable increase
in quality.

## Volatile Acidity

The data description states that high levels of acetic acid can lead to a vinegar
taste.  It seems like this holds true as our lower quality wine have a higher acidity
level.

```{r echo=TRUE, message=FALSE, warning=FALSE, biv_vol}

p1 <- ggplot(wine, aes(rating, volatile.acidity)) +
  geom_jitter(alpha = 0.1) +
  stat_summary(color = 'red') +
  ggtitle("Alcohol vs Volatile Acidity ")

p1

wine %>% group_by(rating) %>%
    summarise(Mean_Volitile_Acdity = mean(volatile.acidity), count = n())

```



## Alcohol

Now I'm going to look at the variables that could have an impact on quality.
It seems like the relationship between alcohol and rating is not lineraly positive,
leading to higher scoring wines generally having a higher alcohol content.  
The mean alcohol content for excellent wines was 11.6%, whereas bad quality wines
had an average content of 10.2%.

```{r echo=FALSE, message=FALSE, warning=FALSE, biv_alcohol}
p1 <- ggplot(wine, aes(quality.fact, alcohol)) +
  geom_jitter(alpha = 0.1) +
  stat_summary(color = 'red') +
  ggtitle("Alcohol vs Quality ")


p2 <- ggplot(wine, aes(rating, alcohol)) +
  geom_boxplot(alpha = 0.3, col = 'black') + 
  ggtitle("Alcohol vs Rating ")

grid.arrange(p1, p2)

  wine %>% group_by(rating) %>%
    summarise(mean(alcohol), median(alcohol))

```




# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

- I've explored the relationship between alcohol content and wine rating, it seems 
that higher scoring wines tended to have a higher alcohol content.

- Density seems to decrease with alcohol percentage

- Density increases as residual sugar levels increase

- Lower quality wines seem to have a higher acetic acid content.


### What was the strongest relationship you found?
According to our correlation plot, density was strongly correlated with sugar
and alcohol content levels.

- Alcohol vs Wine QUality
- Alcohol vs Density
- Sugar vs Density


# Multivariate Plots Section

## Free sulphur

According to our correlation plot - free sulfur and quality are not correlated. 
Perhaps the if we take a look at noticeable amounts of free sulfur they
can correlate with quality levels. 
Next up I compared the total sulfur to quality correlations.


```{r echo=FALSE, free_sulphur}
wine_sulfur <- cor(wine$free.sulfur.dioxide, wine$quality)

wine_not_noticeable_sulfur <- 
  cor(subset(wine, free.sulfur.dioxide.cat != "noticeable" )$quality, 
    subset(wine, free.sulfur.dioxide.cat != "noticeable" )$total.sulfur.dioxide)

wine_noticeable_sulfur <- 
  cor(subset(wine, free.sulfur.dioxide.cat == "noticeable" )$quality, 
    subset(wine, free.sulfur.dioxide.cat == "noticeable" )$total.sulfur.dioxide)

paste("Full wine quality/ sulfur correlation:", 
      round(wine_sulfur,digits=4))

paste("Not noticable sulfur / quality correlation:", 
      round(wine_not_noticeable_sulfur,digits=4))

paste("Noteiceable sulfur / quality correlation:", 
      round(wine_noticeable_sulfur,digits=4))

```

So when we have a noticeable amount of sulfur in the wine there is a stronger 
negative correlation between total sulfur and quality.


```{r echo=FALSE, free_sulphur2}

ggplot( aes( x= rating, y = total.sulfur.dioxide), data = wine) +
  geom_jitter(alpha= 0.3, size = 0.9, aes(color = free.sulfur.dioxide.cat)) +
  ggtitle("Rating vs Total Sulfur Dioxide")


```

Total SO2 content is made up of free and bound sulfur content, it's interesting
to see how some wines have a higher total sulfur content but it is still not
noticeable as I assume free sulfur dioxide levels remain below 50.

Lets explore how the noticeable free sulfur dioxide levels relate to alcohol and 
quality.


```{r echo=FALSE, Multivariate_Plots}


ggplot(wine, aes(quality.fact, alcohol, fill = free.sulfur.dioxide.cat)) +
    geom_boxplot() +
    ggtitle("Alcohol vs Quality")

wine %>% group_by(quality.fact, free.sulfur.dioxide.cat) %>%
    summarise(mean(alcohol), count = n())


```
It seems almost consistently wines with not noticeable sulfur levels have a higher
alcohol content. A higher alcohol content is a key variable in leading to the quality
of wine.
Therefore if you're looking to avoid sulfur, pick a wine with a higher alcohol 
content!

## Alcohol vs Volatile Acidity

We know wines with a higher alcohol percentage and lower acetic acid level tend to be 
a higher quality.
Lets visualise this by looking at the "bad" and "excellent" wines in our data.

```{r echo=FALSE, Multivariate_Plots2}

ggplot(subset(wine, rating != "average"), aes(alcohol, volatile.acidity, color = rating)) +
  geom_point() +
  ggtitle("Alcohol vs Volatile Acidity")

```

Straight away we can see that this holds true, most of our higher quality wines
are in the bottom right of the graph showing they have a higher alcohol content
and lower acetic acid level.


# Multivariate Analysis

Three key relationships were discovered here, firstly we looked at how the 
levels of SO2 in wine can have different total amounts yet still be undectable due
to unnoticeable levels of SO2.  Once SO2 levels were noticeabel there is a slight 
correlation between total SO2 levels and quality.

Next, we explored how white wines with a higher alcohol content generally have a
lower level of total SO2.

Finally we explored two characteristics in white wine and how that higher quality
white wines generally have lower volitile acid levels as well as a higher alcohol 
percentage.

------
  
# Final Plots and Summary

## Plot one
```{r echo=FALSE, Plot_One}
  p1 = ggplot(wine, aes(quality)) +
  geom_bar(fill="steelblue") +
  scale_x_continuous(breaks=seq(3,9,1), limits = c(2,10)) +
  xlab("Quality Score") +
  ylab("Number of Wines")

  p2 = ggplot(wine, aes(rating)) +
    geom_bar(fill = "steelblue") +
    xlab("Wine Rating") +
    ylab("Number of Wines")

  grid.arrange(p1, p2, ncol= 2, top=textGrob("Distribution of Wine Quality",gp=gpar(fontsize=20,font=2)))

```


### Description One

The first final plot I've chosen highlights the distribution of wines within the
dataset.  It's important to not there was not a balanced distribution of bad, 
excellent and average wines.  This begs the question, is this the distribution
for wines generally, or just for this dataset?
The data also contained no entries for wines have a score of 1, 2 or 10 - I can see 
how this is possible as quality scores are defined as the mean from 3 expert
opinions, making it difficult to achieve these scores.


## Plot Two
```{r echo=FALSE, Plot_Two}
  ggplot(wine, aes(rating, alcohol, fill = rating)) + 
  geom_boxplot() +
  geom_smooth(method = "lm", se=FALSE, color="red", aes(group=1)) + 
  scale_fill_brewer() + 
  ggtitle('Wine Quality vs. Alcohol Level') + 
  ylab('Alcohol Level (% by Volume)') +
  xlab('Wine Quality') 

```

### Description Two

My next plot shows how a characteristic of excellent wine is having a higher alcohol 
percentage.  This was the strongest correlation I found within my data of 0.44,
however this is not particularly strong so should not be used as the primary indicator
in white wine quality.


## Plot Three
```{r echo=FALSE, Plot3}

ggplot(subset(wine, rating != "average"), aes(alcohol, volatile.acidity, color = rating)) +
  geom_point() +
  ggtitle("Alcohol vs Volatile Acidity") +
  xlab('Alcohol Level (% by Volume)') +
  ylab('Volatile Acidity') 

```

### Description Three

During the multivariant analysis I discovered that higher quality wines tended
to have a higher alcohol content as well as a lower volatile acidity content.
It's possible that his has a synergistic effect on the taste of the wine as a higher
alcohol percentage could also mask the acidity taste from wine.

------

# Reflection

I feel that wine quality is decided by many independent factors, it is very 
difficult to distinguish wine quality just by combinding of two factors. 
Also, a large number of factors were not present in this data that could have 
accounted for a change in quality such as wine price, brand or grape type.
This dataset had limited data on excellent and poorer qualities of wines, 
making it difficult to produce a representative analysis. 
Findings from this analysis can only be generalised to the Portuguese "Vinho Verde" 
wine, it's very likely that different types of white wine will have varyingly
different factors that conttribute towards quality.

# References
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt
